Comparison of particle trajectories and collision operators for collisional transport in nonaxisymmetric plasmas

### 2014-01-06 doc/sfincsPaper/sfincsPaper.tex
Added documentation

\newcommand{\red}[1]{\textcolor{red}{#1}}
\newcommand{\vect}[1]{\mbox{\boldmath $#1$}}
\newcommand{\erf}{\mathrm{erf}}
\newcommand{\PS}{Pfirsch-Schl\"{u}ter~}
\newcommand{\ExB}{\vect{E}\times\vect{B}}
\newcommand{\Cnl}{C_{\mathrm{nl}a}}
\newcommand{\vma}{\vect{v}_{\mathrm{m}a}}
\newcommand{\vm}{\vect{v}_{\mathrm{m}}}
\newcommand{\vda}{\vect{v}_{\mathrm{d}a}}
\newcommand{\vE}{\vect{v}_{E}}
\newcommand{\vdao}{\vect{v}_{\mathrm{d}a0}}
\newcommand{\vdo}{\vect{v}_{\mathrm{d}0}}
\newcommand{\vEo}{\vect{v}_{E0}}
\newcommand{\vpar}{v_{||}}
\newcommand{\vperp}{v_{\bot}}
\newcommand{\rhop}{\rho_{\theta}}
\newcommand{\rT}{r_{T}}
\newcommand{\Ti}{T_{\mathrm{i}}}
\newcommand{\rTi}{r_{T\mathrm{i}}}
\newcommand{\rTe}{r_{T\mathrm{e}}}
\newcommand{\reta}{r_{\eta}}
\newcommand{\rn}{r_{n}}
\newcommand{\rperp}{r_\bot}
\newcommand{\fM}{f_{\mathrm{M}}}
\newcommand{\fMa}{f_{\mathrm{M}a}}
\newcommand{\fMi}{f_{\mathrm{M}i}}
\newcommand{\fMe}{f_{\mathrm{Me}}}
\newcommand{\fe}{f_{\mathrm{e}}}
\newcommand{\Te}{T_{\mathrm{e}}}
\newcommand{\vth}{v_{\mathrm{th}}}
\newcommand{\nuii}{\nu_{ii}}
\newcommand{\nuPrime}{\hat{\nu}}
\newcommand{\gradpar}{\nabla_{||}}
\newcommand{\kpar}{k_{||}}
\newcommand{\psia}{\psi_0}
\newcommand{\psiaHat}{\hat{\psi}_0}
\newcommand{\todo}[1]{\textcolor{red}{To do: #1}}
\newcommand{\changed}[1]{\textcolor{green}{#1}}
\newcommand{\sgn}{\mathrm{sgn}}

@@

\author{M Landreman}
\email[]{mattland@umd.edu}
%\homepage[]{Your web page}
%\thanks{}
\affiliation{Institute for Research in Electronics and Applied Physics, University of Maryland, College Park, MD, 20742, USA}
\author{H M Smith}
\author{P Helander}
\affiliation{Max-Planck-Institut f\"{u}r Plasmaphysik, 17491 Greifswald, Germany}

@@

Neoclassical phenomena are computed numerically for the LHD and W7-X stellarators using several variants of the drift-kinetic
equation, differing in the terms involving the radial electric field. It is found that
for electric fields below roughly one third of the resonant value, the different formulations give
nearly identical results, demonstrating the incompressible ExB-drift approximation
is quite accurate in this regime. 
However, near the electric field resonance,
the models yield substantially different results.
We also compare results for various collision operators.
In many cases it is found to be important that collisions conserve momentum,
while differences between the full linearized Fokker-Planck operator
and a simpler momentum-conserving model operator are often relatively minor.

@@

One important difference between axisymmetric and nonaxisymmetric plasmas is that
neoclassical effects in the latter are much more sensitive to small values of the radial
electric field $E_r$. In axisymmetric plasmas, in order for the radial electric to modify the collisional ion heat flux
and other neoclassical phenomena,
the poloidal ion Mach number $(B / B_{pol}) |\vect{v}_E|/v_i$ must approach $\sim 1$,
since an $E_r$ of corresponding magnitude is required to modify the trapped region of phase space\cite{GrishaNeo}.
Here, $B$ is the magnetic field magnitude, $B_{pol}$ is the poloidal magnetic field,
$\vect{v}_E$ is the $\vect{E}\times\vect{B}$ drift, and $v_i=\sqrt{2 T_i/m_i}$ is the ion thermal speed.
However, in nonaxisymmetric plasmas, 
a much smaller value of $E_r$ can modify the collisional fluxes \cite{Galeev, HoKulsrud, BeidlerBigBenchmarking}.
The reason is that helically trapped particles experience a secular radial magnetic drift, and
whichever process first interrupts this radial motion will thereby determine the step size for radial diffusion.
When $E_r=0$, the radial magnetic drift is interrupted by collisions, which cause
the particle to gain parallel momentum and de-trap.
But if $E_r$ is sufficient for the poloidal $\vect{E}\times\vect{B}$
precession frequency to exceed the effective collisional detrapping rate, $\vect{E}\times\vect{B}$ precession
begins to carry helically trapped particles onto untrapped trajectories, thereby limiting the radial step size and transport.
This transition from collisional ($1/\nu$-regime) to $E_r$-limited ($\sqrt{\nu}$-regime) transport
typically occurs at values of $E_r$ for which the poloidal Mach number is still $\ll 1$, due to the low collisionality
in typical experiments. For this reason, stellarator transport is sensitive to small values of $E_r$.

However, including the physics of $\vect{E}\times\vect{B}$ precession in a $\delta f$ drift-kinetic equation (or code to
solve such an equation)
is complicated by several issues.
First, if a rigorous expansion in $\rho_* \ll 1$ is employed, 
$\vect{E}\times\vect{B}$ is formally excluded
when the usual drift ordering $\vect{v}_E \sim \rho_* v_i$ is used,
but the high-flow ordering $\vect{v}_E \sim \rho_*^0 v_i$ is
not a useful ordering either, since it leads to contradictions in a general nonaxisymmetric field \cite{HelanderRotation, SugamaHighFlow}.
Here, $\rho_* = \rho/L$ where $\rho$ is the ion gyroradius and $L$ is a typical
macroscopic scale length.
Second, if the $\vect{v}_{E}$ poloidal precession term
is included in a radially local, time-independent kinetic equation
for $\delta f$ (the departure of the distribution function from a Maxwellian),
unphysical constraints are placed on the distribution function,
as we will prove in section \ref{sec:sources}.
These constraints only appear when $E_r \ne 0$, meaning a small but nonzero
$E_r$ is a singular perturbation of the $E_r=0$ case.
These unphysical behaviors have been elininated in previous codes\cite{BeidlerBigBenchmarking}
by making the ad-hoc replacement $1/B^2 \to 1/\left< B^2 \right>$ (where $\left< \ldots \right>$ denotes
a flux surface average) in the $\vect{E}\times\vect{B}$ drift.
At the same time, variation in the particles' energy and pitch angle associated with $E_r$ is neglected.
These replacements and omissions are chosen so as to restore the variational form of the kinetic equation  \cite{DKES1, DKES2}.
These changes to the kinetic equation may be called the ``incompressible-$\vect{E}\times\vect{B}$-drift''
approximation \cite{Beidler}.
Some investigations have indicated that the incompressible-$\vect{E}\times\vect{B}$-drift
approximation may be reasonably accurate for small $E_r$ but a poor approximation
for larger $E_r$ \cite{Beidler, meMonoenergetic}.
This issue of which collisionless terms to include in the kinetic equation
is effectively a choice between particle trajectories, since the collisionless guiding center trajectories
are equivalent to the characteristic curves of the drift-kinetic equation.

\begin{comment}
If the typical drift ordering $\vect{v}_E \sim \rho_* v_i$ is used,
with the usual ordering of the collision frequency $\nu$ as $\sim \rho_*^0 v_i/L$,
and a rigorous ordering of terms in $\rho_*$ is made, collisionless detrapping is formally
excluded. Here, $\rho_* = \rho/L$ where $\rho$ is the ion gyroradius and $L$ is a typical
macroscopic scale length.  Collisionless detrapping cannot be included by
the alternative ordering $\vect{v}_E \sim v_i$,
since this ordering leads to contradictions in a general nonaxisymmetric field \cite{HelanderRotation, SugamaHighFlow}.
In many stellarator neoclassical calculations and codes, then, the $\rho_*$ ordering is not rigidly followed,
with the $v_E$ poloidal precession term

To include the physics of collisionless detrapping, $\vect{E}\times\vect{B}$-drift terms
must be retained in the kinetic equation even though these terms are smaller
than other terms in the equation according to the $\rho_* \ll 1$ expansion.
At the same time, the ad-hoc replacement $1/B^2 \to 1/\left< B^2 \right>$ (where $\left< \ldots \right>$ denotes
a flux surface average) is sometimes made in the $\vect{E}\times\vect{B}$ drift \cite{DKES1, DKES2},
since this replacement happens to restore the variational form of the kinetic equation.
This replacement may be called the ``incompressible-$\vect{E}\times\vect{B}$-drift''
approximation \cite{Beidler}.
...
You can think of these various terms as related to effective particle trajectories.
\end{comment}

Another limitation of many past stellarator neoclassical calculations
is that they are often performed with simplified models for collisions \todo{Cite other stellarator codes here}.
The linearized Fokker-Planck collision operator -- the most accurate linear operator available -- 
has been implemented in a variety of 
tokamak neoclassical codes \cite{Sauter0, Sauter, WongChan, NEOFP, speedGrids}.
However, due to the numerical challenge of the extra toroidal dimension in stellarators, 
many stellarator neoclassical codes retain only pitch-angle scattering collisions,
so coupling in the energy dimension is eliminated.
The pitch-angle scattering operator
lacks the momentum conservation property of the Fokker-Planck operator,
which is known to be important in many situations \cite{Briesemeister}.
Several techniques have been devised and implemented \cite{Taguchi2, SugamaNishimura, Maassberg, PENTA}
to effectively restore momentum conservation by post-processing
the transport coefficients obtained with a pure pitch-angle scattering operator,
but these methods will not exactly reproduce calculations with the full linearized Fokker-Planck operator.
The NEO-2 code has implemented the full linearized Fokker-Planck operator for
stellarator geometry \cite{Kernbichler1, Kernbichler2}, but using a field-line-tracing method
which makes it difficult to add the important effect of poloidal $\vect{E}\times\vect{B}$ precession.

Here, we describe a new stellarator neoclassical code
SFINCS (the Stellarator Fokker-Planck Iterative Neoclassical Conservative Solver)
that can be used to explore the aforementioned issues,
comparing various models for effective particle trajectories and collisions.
The code solves the 4D drift-kinetic equation for the distribution function,
retaining coupling in 2 spatial independent variables (toroidal and poloidal angle)
and 2 velocity independent variables (speed and pitch angle),
but neglecting radial coupling.
(For comparison, DKES \cite{DKES1, DKES2} is 3D since energy coupling is neglected, while FORTEC-3D \cite{FORTEC} is 5D since
radial coupling is retained.)
General nonaxisymmetric nested flux surface geometry is allowed,
one or more species may be included, and several models for collisions are available,
including the full inter-species linearized Fokker-Planck operator.
The incompressible-$\vect{E}\times\vect{B}$-drift trajectories are implemented,
as are several other options for trajectories that include the true $\vect{E}\times\vect{B}$ drift.
As we shall demonstrate, retaining the true form of the $\vect{E}\times\vect{B}$
drift comes at a cost, requiring sources/sinks in the kinetic equation
in order for the solutions to be well behaved.
While all of the various options for the particle trajectories have disadvantages,
SFINCS allows the three options to be compared.
As we will show in several calculations for the LHD and W7-X stellarators,
in many experimentally relevant cases,
the transport matrix elements are nearly identical for the various choices of particle trajectories.
However, differences between the trajectory models emerge when the radial electric field
grows comparable to the ``resonant'' value.

In the following section, we motivate the form of the kinetic equation solved by SFINCS,
and detail the three models for particle trajectories that will be compared.
Section \ref{sec:collisionOperators} describes several available models for collisions.
For several of the particle trajectory models,
additional sources/sinks and constraints must be included in the system of equations
for the equations to be well posed and for the solutions to be well behaved.
These issues are explored in section \ref{sec:sources}.
Details of the numerical implementation are given in section \ref{sec:numerics}.
Some of the numerical results presented are given in terms of a transport
matrix, which is defined in section \ref{sec:transportMatrix}.
The numerical results are presented in sections \ref{sec:ErComparison} and \ref{sec:collisionComparison},
in which we discuss the transport matrix elements for the geometries of the LHD and W7-X
stellarators, comparing a variety of assumptions about the particle trajectories and collision operator.
In section \ref{sec:conclusions} we discuss the results and conclude.

\todo{cite EUTERPE and FORTEC-3D}

@@

We begin with the drift-kinetic equation (19) of Ref. \cite{Hazeltine}.
The standard drift ordering is applied at first: $\rho_* \ll 1$ where $\rho_* = \rho/L$,
$\vE/v_a \sim \rho_*$, $\partial/\partial t \sim \rho_*^2 v_a/L$, and $\nu_a \sim v_a/L$,
where $L$ is a typical scale length.
We expand the distribution function as $f_a = f_{a0} + f_{a1} + \ldots$.
The leading order distribution function $f_{a0}$ is taken to be a Maxwellian
that is constant on flux surfaces when expressed in terms of total energy
$W_a = v^2/2+Z_a e \Phi/m_a$:
\begin{equation}
f_{a0} = \eta_a(\psi) \left[ \frac{m_a}{2\pi T_a(\psi)}\right]^{3/2} \exp\left( -\frac{m_a W_a}{T_a(\psi)}\right).
\label{eq:Maxwellian}
\end{equation}
The mean flow of this Maxwellian is taken to be zero since, as argued in Refs. \cite{HelanderRotation, SugamaHighFlow}, sonic
flows are not permitted in a general stellarator.
Taking $f_{a1}/f_{a0}\sim\rho_*$, the terms of order $\sim\rho_* (v/L)f_{a0}$ in (19) of Ref. \cite{Hazeltine} are then
\begin{equation}
\vpar \vect{b} \cdot\left(\nabla f_{a1}\right)_{W_a,\mu}
-C_{a}
=-\left(\vma + \vE\right)\cdot\nabla\psi \left(\frac{\partial f_{a0}}{\partial \psi}\right)_{W_a}
+\frac{Z_a e}{m_a c}\vpar\vect{b}\cdot\frac{\partial\vect{A}}{\partial t} \frac{\partial f_{a0}}{\partial W_a}
\label{eq:kinetic1}
\end{equation}
where
\begin{equation}
v_{ma}\cdot\nabla\psi = \frac{m_a c v_{||}^2}{Z_a e B} \vect{b}\times (\vect{b}\cdot\nabla\vect{b})\cdot\nabla\psi
+ \frac{m_a c v_{\perp}^2}{2 Z_a e B^2} \vect{b}\times \nabla B \cdot\nabla\psi
=\frac{m_a c}{2 Z_a e B^2}\left(v_{||}^2 + \frac{v_{\bot}^2}{2}\right) \vect{b}\times \nabla B \cdot\nabla\psi
\end{equation}
(exactly true for any $\beta$) and $\vE=(c/B^2)\vect{B}\times\nabla\Phi$.
Here, $C_a$ is the collision term for species $a$, linearized about the Maxwellians (\ref{eq:Maxwellian}).
Subscripts on gradients and partial derivatives indicate the quantities held fixed,
and $\mu=v^2/(2B)$ is the magnetic moment.

Unfortunately, (\ref{eq:kinetic1}) does not contain the physics of $\vect{E}\times\vect{B}$ precession,
since the characteristic curves of this equation correspond only to motion along the magnetic field lines.
Consequently, important transport regimes such as the $\sqrt{\nu}$ regime cannot be obtained
using (\ref{eq:kinetic1}).
To retain $\vect{E}\times\vect{B}$ precession, we also keep the term $(\vE+\vma)\cdot\nabla f_{a1}$
in (\ref{eq:kinetic1}), even though according to the formal ordering it should
appear at next order.  A similar step is made in other stellarator neoclassical
calculations \cite{DKES1, DKES2}

As shown in Appendix C of Ref. \cite{usPedestal}, we may choose the gauge
for the electromagnetic potentials such that $-c^{-1}\vect{b}\cdot\partial\vect{A}/\partial t = \left<E_{||} B\right> B/\left<B^2\right>$
on the right-hand side of (\ref{eq:kinetic1}).
Here, angle brackets denote a flux surface average:
\begin{equation}
\label{eq:FSA}
\left\langle \ldots \right\rangle
=\frac{1}{{V}'}
\int_0^{2\pi }d\theta
\int_0^{2\pi }d\zeta
{\frac{(\ldots)}{\vect{B}\cdot \nabla \zeta }}
\end{equation}
where
${V}'=\int_0^{2\pi } d\theta \int_0^{2\pi } d\zeta / \vect{B}\cdot \nabla \zeta $.
Thus, (\ref{eq:kinetic1}) becomes
\begin{equation}
\left(\vpar \vect{b} + \vE + \vma\right) \cdot\left(\nabla f_{a1}\right)_{W_a,\mu}
-C_{a}
=-\left(\vma + \vE\right)\cdot\nabla\psi \left(\frac{\partial f_{a0}}{\partial \psi}\right)_{W_a}
+\frac{Z_a e}{T_a}\vpar\frac{B\left<E_{||} B\right>}{\left< B^2\right>} f_{a0}.
\label{eq:kinetic2}
\end{equation}
Even if the radial electric field is considered an input, this form of the kinetic equation remains
nonlinear in the unknowns since the $\nabla f_{a1}$ term depends on the variation of $\Phi$
on a flux surface, and this variation is an unknown like $f_{a1}$.

>>>>

\newcommand{\red}[1]{\textcolor{red}{#1}}
\newcommand{\vect}[1]{\mbox{\boldmath $#1$}}
\newcommand{\erf}{\mathrm{erf}}
\newcommand{\PS}{Pfirsch-Schl\"{u}ter~}
\newcommand{\ExB}{\vect{E}\times\vect{B}}
\newcommand{\Cnl}{C_{\mathrm{nl}a}}
\newcommand{\vma}{\vect{v}_{\mathrm{m}a}}
\newcommand{\vm}{\vect{v}_{\mathrm{m}}}
\newcommand{\vda}{\vect{v}_{\mathrm{d}a}}
\newcommand{\vE}{\vect{v}_{E}}
\newcommand{\vdao}{\vect{v}_{\mathrm{d}a0}}
\newcommand{\vdo}{\vect{v}_{\mathrm{d}0}}
\newcommand{\vEo}{\vect{v}_{E0}}
\newcommand{\vpar}{v_{||}}
\newcommand{\vperp}{v_{\bot}}
\newcommand{\rhop}{\rho_{\theta}}
\newcommand{\rT}{r_{T}}
\newcommand{\vi}{v_{\mathrm{i}}}
\newcommand{\me}{m_{\mathrm{e}}}
\newcommand{\mi}{m_{\mathrm{i}}}
\newcommand{\Ti}{T_{\mathrm{i}}}
\newcommand{\rTi}{r_{T\mathrm{i}}}
\newcommand{\rTe}{r_{T\mathrm{e}}}
\newcommand{\reta}{r_{\eta}}
\newcommand{\rn}{r_{n}}
\newcommand{\rperp}{r_\bot}
\newcommand{\fM}{f_{\mathrm{M}}}
\newcommand{\fMa}{f_{\mathrm{M}a}}
\newcommand{\fMi}{f_{\mathrm{M}i}}
\newcommand{\fMe}{f_{\mathrm{Me}}}
\newcommand{\fe}{f_{\mathrm{e}}}
\newcommand{\Te}{T_{\mathrm{e}}}
\newcommand{\vth}{v_{\mathrm{th}}}
\newcommand{\nuii}{\nu_{\mathrm{ii}}}
\newcommand{\nuPrime}{\hat{\nu}}
\newcommand{\gradpar}{\nabla_{||}}
\newcommand{\kpar}{k_{||}}
\newcommand{\psia}{\psi_0}
\newcommand{\psiaHat}{\hat{\psi}_0}
\newcommand{\todo}[1]{\textcolor{red}{To do: #1}}
\newcommand{\changed}[1]{\textcolor{green}{#1}}
\newcommand{\sgn}{\mathrm{sgn}}
\newcommand{\Erres}{E_r^{\mathrm{res}}}
\newcommand{\Bpol}{B_{\mathrm{pol}}}
\newcommand{\Sap}{S_{a\mathrm{p}}}
\newcommand{\Sah}{S_{a\mathrm{h}}}

@@

\author{M Landreman}
\email[]{mattland@umd.edu}
%\homepage[]{Your web page}
%\thanks{}
\affiliation{Institute for Research in Electronics and Applied Physics, University of Maryland, College Park, MD, 20742, USA}
\affiliation{Plasma Science and Fusion Center, Massachusetts Institute of Technology, Cambridge, MA, 02139, USA}
\author{H M Smith}
\author{P Helander}
\affiliation{Max-Planck-Institut f\"{u}r Plasmaphysik, 17491 Greifswald, Germany}

@@

In this work, we test the validity of several common simplifying assumptions used in numerical neoclassical calculations
for nonaxisymmetric plasmas.
First, neoclassical phenomena are computed for the LHD and W7-X stellarators using several versions of the drift-kinetic
equation, 
including the commonly used incompressible-$\vect{E}\times\vect{B}$-drift approximation
and two other variants,
corresponding to different effective particle trajectories.
It is found that
for electric fields below roughly one third of the resonant value, the different formulations give
nearly identical results, demonstrating the incompressible $\vect{E}\times\vect{B}$-drift approximation
is quite accurate in this regime. 
However, near the electric field resonance,
the models yield substantially different results.
We also compare results for various collision operators, including the full linearized Fokker-Planck operator.
At low collisionality, the radial transport driven by radial
gradients is nearly identical for the different operators,
while in other cases it is found to be important that collisions conserve momentum.
%At high collisionality, it seems necessary to use the full Fokker-Planck operator to obtain all transport coefficients correctly.

@@

One important difference between axisymmetric and nonaxisymmetric plasmas is
that neoclassical effects in the latter are more sensitive to small values of the radial
electric field $E_r$. In axisymmetric plasmas, in order for the radial electric to modify the collisional ion heat flux
and other neoclassical phenomena,
the poloidal ion Mach number $(B / \Bpol) |\vect{v}_E|/\vi$ must approach $\sim 1$,
since an $E_r$ of corresponding magnitude is required to modify the trapped region of phase space\cite{GrishaNeo}.
Here, $B$ is the magnetic field magnitude, $\Bpol$ is the poloidal magnetic field,
$\vect{v}_E$ is the $\vect{E}\times\vect{B}$ drift, and $\vi=\sqrt{2 \Ti / \mi}$ is the ion thermal speed.
However, in nonaxisymmetric plasmas, 
a much smaller value of $E_r$ can modify the collisional fluxes \cite{Galeev, HoKulsrud, BeidlerBigBenchmarking}.
The reason is that helically trapped particles experience a secular radial magnetic drift, and
whichever process first interrupts this radial motion will thereby determine the step size for radial diffusion.
When $E_r=0$, the radial magnetic drift is interrupted by collisions, which cause
the particle to gain parallel momentum and de-trap.
But if $E_r$ is sufficient for the poloidal $\vect{E}\times\vect{B}$
precession frequency to exceed the effective collisional detrapping rate, $\vect{E}\times\vect{B}$ precession
begins to carry helically trapped particles onto untrapped trajectories, 
and also confines the trapped orbits by convecting them (usually poloidally) around the torus,
thereby limiting the radial step size and transport.
This transition from collisional ($1/\nu$-regime) to $E_r$-limited ($\sqrt{\nu}$-regime) transport
typically occurs at values of $E_r$ for which the poloidal Mach number is still $\ll 1$, due to the low collisionality
in typical experiments. (Here, $\nu$ denotes a collision frequency.)
For this reason, stellarator transport at low collisionality is sensitive to small values of $E_r$.
A variety of codes have been developed to compute 
these neoclassical effects in stellarators \cite{DKES1, DKES2, GSRAKE, Tribaldos2001, PENTA, Kernbichler1, FORTEC_PFR, FORTEC_CPC, JOSE}.

However, including the physics of $\vect{E}\times\vect{B}$ precession in a $\delta\! f$ drift-kinetic equation (or code to
solve such an equation)
is complicated by several issues.
First, if a rigorous expansion in $\rho_* \ll 1$ is employed, 
$\vect{E}\times\vect{B}$ precession is formally excluded
when the usual drift ordering $\vect{v}_E \sim \rho_* \vi$ is used,
but the high-flow ordering $\vect{v}_E \sim \vi$ is
not a useful ordering either, since it leads to contradictions in a general nonaxisymmetric field \cite{HelanderRotation, SugamaHighFlow}.
Here, $\rho_* = \rho/L$ where $\rho$ is the ion gyroradius and $L$ is a typical
macroscopic scale length.
Second, if the $\vect{v}_{E}$ poloidal precession term
is included in a radially local, time-independent kinetic equation
for $\delta\! f$ (the departure of the distribution function from a Maxwellian),
unphysical constraints are placed on the distribution function,
as we will prove in section \ref{sec:sources}
by considering appropriate moments of the kinetic equation.
These constraints only appear when $E_r \ne 0$, meaning a small but nonzero
$E_r$ is a singular perturbation of the $E_r=0$ case.
These unphysical behaviors have been eliminated in previous codes\cite{BeidlerBigBenchmarking}
by making the ad-hoc replacement $1/B^2 \to 1/\left< B^2 \right>$ (where $\left< \ldots \right>$ denotes
a flux surface average) in the $\vect{E}\times\vect{B}$ drift.
At the same time, variation in the particles' energy and pitch angle associated with $E_r$ is neglected.
These replacements and omissions are chosen so as to restore the variational form of the kinetic equation  \cite{DKES1, DKES2}.
These changes to the kinetic equation may be called the ``incompressible-$\vect{E}\times\vect{B}$-drift''
approximation \cite{Beidler}.
Some investigations have indicated that the incompressible-$\vect{E}\times\vect{B}$-drift
approximation may be reasonably accurate for small $E_r$ but a poor approximation
for larger $E_r$ \cite{Beidler, meMonoenergetic}.
This issue of which collisionless terms to include in the kinetic equation
is effectively a choice between particle trajectories, since the collisionless guiding center trajectories
are equivalent to the characteristic curves of the drift-kinetic equation.

Another limitation of many past stellarator neoclassical calculations
is that they are often performed with simplified models for collisions.
The linearized Fokker-Planck collision operator -- the most accurate linear operator available -- 
has been implemented in a variety of 
tokamak neoclassical codes \cite{Sauter0, Sauter, WongChan, NEOFP, speedGrids}.
However, due to the numerical challenge of the extra toroidal dimension in stellarators, 
many stellarator neoclassical codes retain only pitch-angle scattering collisions,
so coupling in the energy dimension is eliminated.
The pitch-angle scattering operator
lacks the momentum conservation property of the Fokker-Planck operator,
which is known to be important in many situations \cite{Briesemeister}.
Several techniques have been devised and implemented \cite{Taguchi2, SugamaNishimura, Maassberg, PENTA}
to effectively restore momentum conservation by post-processing
the transport coefficients obtained with a pure pitch-angle scattering operator,
but these methods will not exactly reproduce calculations with the full linearized Fokker-Planck operator.
The NEO-2 code has implemented the full linearized Fokker-Planck operator for
stellarator geometry \cite{Kernbichler1}, but using a field-line-tracing method
which makes it difficult to add the important effect of poloidal $\vect{E}\times\vect{B}$ precession.

Here, we describe a new stellarator neoclassical code
SFINCS (the Stellarator Fokker-Planck Iterative Neoclassical Conservative Solver)
that can be used to explore the aforementioned issues,
comparing various models for effective particle trajectories and collisions.
Although we use the terminology of ``effective trajectories,'' the code
uses continuum rather than Monte Carlo algorithms.
The code solves the 4D drift-kinetic equation for the distribution function,
retaining coupling in 2 spatial independent variables (toroidal and poloidal angle)
and 2 velocity independent variables (speed and pitch angle),
but neglecting radial coupling.
(For comparison, DKES \cite{DKES1, DKES2} is 3D since energy coupling is neglected, 
while FORTEC-3D \cite{FORTEC_PFR, FORTEC_CPC} is 5D since
radial coupling is retained.)
General nonaxisymmetric nested flux surface geometry is allowed,
one or more species may be included, and several models for collisions are available,
including the full inter-species linearized Fokker-Planck operator.
The incompressible-$\vect{E}\times\vect{B}$-drift trajectories are implemented,
as are several other options for trajectories that include the true $\vect{E}\times\vect{B}$ drift.
As we shall demonstrate, retaining the true form of the $\vect{E}\times\vect{B}$
drift comes at a cost, requiring sources/sinks in the kinetic equation
in order for the solutions to be well behaved.
While all of the various options for the particle trajectories have disadvantages,
SFINCS allows the three options to be compared.
As we will show in several calculations for the LHD and W7-X stellarators,
in many experimentally relevant cases,
the transport matrix elements are nearly identical for the various choices of particle trajectories.
However, differences between the trajectory models emerge when the radial electric field
grows comparable to the ``resonant'' value.

In the following section, we motivate the form of the kinetic equation solved by SFINCS,
and detail the three models for particle trajectories that will be compared.
For several of the particle trajectory models,
additional sources/sinks and constraints must be included in the system of equations
for the equations to be well posed and for the solutions to be well behaved.
These issues are explored in section \ref{sec:sources}.
Details of the numerical implementation are given in section \ref{sec:numerics}.
Some of the numerical results presented are given in terms of a transport
matrix, which is defined in section \ref{sec:transportMatrix}.
The numerical results are presented in sections \ref{sec:ErComparison} and \ref{sec:collisionComparison},
in which we discuss the transport matrix elements for the geometries of the LHD and W7-X
stellarators, comparing a variety of assumptions about the particle trajectories and collision operator.
In section \ref{sec:conclusions} we discuss the results and conclude.

@@

We begin with the drift-kinetic equation (19) of Ref. \cite{Hazeltine}.
The standard drift ordering is applied at first: $\rho_{*a} \ll 1$ where $\rho_{*a} = \rho_a/L$,
$\vE/v_a \sim \rho_{*a}$, $\partial/\partial t \sim \rho_{*a}^2 v_a/L$, and $\nu_a \sim v_a/L$.
Here, $v_a = \sqrt{2 T_a / m_a}$ is the thermal speed of species $a$, $T_a$ is the temperature,
$m_a$ is the mass, $\rho_a = v_a m_a c/(Z_a e B)$ is the gyroradius, $Z_a$ is the species charge in units
of the proton charge $e$, $c$ is the speed of light, $L$ is a typical scale length,
and $\nu_a$ is a collision frequency.
No expansion in mass ratios or charges is made.
We expand the distribution function as $f_a = f_{a0} + f_{a1} + \ldots$.
The leading order distribution function $f_{a0}$ is taken to be a Maxwellian
that is constant on flux surfaces when expressed in terms of total energy
$W_a = v^2/2+Z_a e \Phi/m_a$:
\begin{equation}
f_{a0} = \eta_a(\psi) \left[ \frac{m_a}{2\pi T_a(\psi)}\right]^{3/2} \exp\left( -\frac{m_a W_a}{T_a(\psi)}\right).
\label{eq:Maxwellian}
\end{equation}
Here, $\Phi$ is the electrostatic potential and $v$ is the speed.
The mean flow of this Maxwellian is taken to be zero since, as argued in Refs. \cite{HelanderRotation, SugamaHighFlow}, sonic
flows are not permitted in a general stellarator.
Taking $f_{a1}/f_{a0}\sim\rho_{*a}$, the terms of order $\sim\rho_{*a} (v/L)f_{a0}$ in (19) of Ref. \cite{Hazeltine} are then
\begin{equation}
\vpar \vect{b} \cdot\left(\nabla f_{a1}\right)_{W_a,\mu}
-C_{a}
=-\left(\vma + \vE\right)\cdot\nabla\psi \left(\frac{\partial f_{a0}}{\partial \psi}\right)_{W_a}
+\frac{Z_a e}{m_a c}\vpar\vect{b}\cdot\frac{\partial\vect{A}}{\partial t} \frac{\partial f_{a0}}{\partial W_a}
\label{eq:kinetic1}
\end{equation}
where the radial magnetic drift is
\begin{equation}
\vma\cdot\nabla\psi = \frac{m_a c v_{||}^2}{Z_a e B} \vect{b}\times (\vect{b}\cdot\nabla\vect{b})\cdot\nabla\psi
+ \frac{m_a c v_{\perp}^2}{2 Z_a e B^2} \vect{b}\times \nabla B \cdot\nabla\psi
=\frac{m_a c}{2 Z_a e B^2}\left(v_{||}^2 + \frac{v_{\bot}^2}{2}\right) \vect{b}\times \nabla B \cdot\nabla\psi
\end{equation}
(exactly true for any $\beta$ in a magnetic equilibrium with isotropic pressure) and the $\vect{E}\times\vect{B}$ drift
is $\vE=(c/B^2)\vect{B}\times\nabla\Phi$.
Here, $\vect{b} = \vect{B}/B$ is the unit vector along the magnetic field, 
$v_{||}$ and $v_{\bot}$ denote the components of velocity parallel and perpendicular to $\vect{B}$,
$2\pi\psi$ is the toroidal flux, $\vect{A}$ is the magnetic vector potential, 
and $C_a$ is the collision term for species $a$, linearized about the Maxwellians (\ref{eq:Maxwellian}).
Subscripts on gradients and partial derivatives indicate the quantities held fixed,
and $\mu=v_{\bot}^2/(2B)$ is the magnetic moment.

Unfortunately, (\ref{eq:kinetic1}) does not contain the physics of $\vect{E}\times\vect{B}$ precession,
since the characteristic curves of this equation correspond only to motion along the magnetic field lines.
Consequently, important transport regimes such as the $\sqrt{\nu}$ regime cannot be obtained
using (\ref{eq:kinetic1}).
To retain $\vect{E}\times\vect{B}$ precession, we also keep the term $(\vE+\vma)\cdot\nabla f_{a1}$
in (\ref{eq:kinetic1}), even though according to the formal ordering it should
appear at next order.  A similar step is made in other stellarator neoclassical
calculations \cite{DKES1, DKES2}.
The mathematical reason why this term is important at low collisionality is that it has different symmetry properties than other, possible larger, terms in (\ref{eq:kinetic1}). For instance, it 
survives if a bounce average is used to annihilate the first term.
(We will not bounce average the kinetic equation here,
but when the collisionality is low, the solution of the full equation becomes
asymptotically close to the solution of the bounce-averaged equation.)

As shown in Appendix C of Ref. \cite{usPedestal}, we may choose the gauge
for the electromagnetic potentials such that $-c^{-1}\vect{b}\cdot\partial\vect{A}/\partial t = \left<E_{||} B\right> B/\left<B^2\right>$
on the right-hand side of (\ref{eq:kinetic1}).
Here, angle brackets denote a flux surface average:
\begin{equation}
\label{eq:FSA}
\left\langle \ldots \right\rangle
=\frac{1}{{V}'}
\int_0^{2\pi }d\theta
\int_0^{2\pi }d\zeta
{\frac{(\ldots)}{\vect{B}\cdot \nabla \zeta }}
\end{equation}
where
${V}'=\int_0^{2\pi } d\theta \int_0^{2\pi } d\zeta / \vect{B}\cdot \nabla \zeta $,
$\theta$ and $\zeta$ are poloidal and toroidal magnetic angles satisfying
\begin{equation}
\vect{B} = \nabla \psi \times \nabla \theta + \iota \nabla\zeta\times\nabla\psi,
\end{equation}
$\iota = 1/q$ is the rotational transform, and $q$ is the safety factor.
Thus, (\ref{eq:kinetic1}) becomes
\begin{equation}
\left(\vpar \vect{b} + \vE + \vma\right) \cdot\left(\nabla f_{a1}\right)_{W_a,\mu}
-C_{a}
=-\left(\vma + \vE\right)\cdot\nabla\psi \left(\frac{\partial f_{a0}}{\partial \psi}\right)_{W_a}
+\frac{Z_a e}{T_a}\vpar\frac{B\left<E_{||} B\right>}{\left< B^2\right>} f_{a0}.
\label{eq:kinetic2}
\end{equation}
Even if the radial electric field is considered an input, this form of the kinetic equation remains
nonlinear in the unknowns since the $\nabla f_{a1}$ term depends on the variation of $\Phi$
on a flux surface, and this variation is an unknown like $f_{a1}$.
